---
title: "LiDAR Structure Analysis"
---

Load in libraries
```{r}

#For LiDAR analysis
.First.sys()
library(here)
library(readxl)
library(terra)
library(ggplot2)
library(stringr)
library(dplyr)
library(lidR)
library(ggimage)
library(magick)
library(rgl)
library(patchwork)
library(tidyr)

#For hierarchical partitioning of LAD PCA
library(dplyr)
library(readxl)
library(vegan)
library(rdacca.hp)
library(ggplot2)
library(ggvenn)
library(ggVennDiagram)
library(tidyverse)
library(ggpattern)

```

PCA of LAD data
```{r}
meta_data <- read_xlsx(here("3_structure_analysis/data/Data_lp.xlsx"), sheet = 2)

#######################################################################################################################
#NOTE: We do not provide the LAD raster file due to sharing restrictions related to the LiDAR data. Please see######### 
#calculate_LAD.Rmd and normalization_gen3.Rmd for how to normalize and calculate LAD from the raw LiDAR data###########
#accessible freely from the Spanish government website of the Centro Nacional de Información Geográfica (CNIG).########
#######################################################################################################################
LAD <- rast(here("3_structure_analysis/data/gen3_La_Palma_LAD.tif"))

buffer <- 50

plots <- vect(meta_data, geom = c("longitude", "latitude"), crs = "EPSG:4326")
plots <- project(plots, "EPSG:32628")
plots <- terra::buffer(plots, buffer)

#writeVector(plots, here("3_structure_analysis/data/plots.shp"), overwrite = TRUE)

LAD_extracted <- extract(LAD, plots)[, -2]

colnames(LAD_extracted)[1] <- "plot"

LAD_extracted <- merge(LAD_extracted, meta_data, by = "plot")

LAD_summary <- LAD_extracted %>%
    group_by(plot) %>%
    summarise(across(c(`level_1_75`:`level_49_25`),
        list(
            mean = ~ mean(.x, na.rm = TRUE),
            sd = ~ sd(.x, na.rm = TRUE)
        ),
        .names = "{.col}_{.fn}"
    ))

LAD_summary <- cbind(LAD_summary, meta_data[, 2:6])

pca_data <- LAD_summary %>%
    select(ends_with("_mean"), ends_with("_sd"))

pca_res <- prcomp(pca_data, scale. = FALSE)

plot_data <- cbind(LAD_summary, pca_res$x[, 1:10])

write_xlsx(plot_data, here("3_structure_analysis/data/PCA_data_structure.xlsx"))

plot_data$color <- str_c(plot_data$age, "_", plot_data$ecosystemtype)

p <- ggplot(plot_data, aes(x = PC1, y = PC2, color = color)) +
    geom_point() +
    stat_ellipse(level = 0.95, size = 1, show.legend = FALSE) +
    scale_color_manual(values = c("#33658A", "#c97704", "#86BBD8", "#FFC15E")) +
    theme_minimal() +
    labs(
        x = paste0("PC1 (", round(summary(pca_res)$importance[2, 1] * 100, 1), "%)"),
        y = paste0("PC2 (", round(summary(pca_res)$importance[2, 2] * 100, 1), "%)"),
        color = "Ecosystem Type"
    ) +
    theme(
        legend.position = "none",
        strip.text = element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line.x = element_line(colour = "black"),
        axis.line.y = element_line(colour = "black"),
        axis.text.x = element_text(size = 12, colour = "black"),
        axis.text.y = element_text(size = 12, colour = "black"),
        axis.title.x = element_text(size = 14, colour = "black"),
        axis.title.y = element_text(size = 14, colour = "black"),
        axis.ticks.length = unit(0.1, "cm"),
        axis.ticks.x = element_line(colour = "black"),
        axis.ticks.y = element_line(colour = "black")
    )

print(p)

```


Hierarchical partitioning of forest structure data
```{r}

data_structure <- read_xlsx(here("3_structure_analysis/data/PCA_data_structure.xlsx"))
data_predictors <- read_xlsx(here("2_diversity_and_community_analysis/data/MvsI_meta_data_29_01_2026.xlsx"))

data_structure_Fa <- data_structure %>%
  filter(ecosystemtype == "Fa")

data_structure_Pi <- data_structure %>%
  filter(ecosystemtype == "Pi")

data_predictors_Fa <- data_predictors %>%
  filter(ecosystemtype == "Fa")

data_predictors_Pi <- data_predictors %>%
  filter(ecosystemtype == "Pi")

resp_Fa <- data.frame(data_structure_Fa$PC1, data_structure_Fa$PC2)

pred_Fa <- data.frame(
  data_predictors_Fa$age,
  data_predictors_Fa$elevation,
  data_predictors_Fa$MAT,
  data_predictors_Fa$MAP
)

resp_Pi <- data.frame(data_structure_Pi$PC1, data_structure_Pi$PC2)

pred_Pi <- data.frame(
  data_predictors_Pi$age,
  data_predictors_Pi$elevation,
  data_predictors_Pi$MAT,
  data_predictors_Pi$MAP
)

hierpar_structure_Fa <- rdacca.hp(
  resp_Fa,
  pred_Fa,
  var.part = F
)

hierpar_structure_Fa

hierpar_structure_Pi <- rdacca.hp(
  resp_Pi,
  pred_Pi,
  var.part = F
)

hierpar_structure_Pi
```

Identify closest plot to the centroid of each group in PCA space
```{r}
plot_centroids <- plot_data %>%
    mutate(row_id = row_number()) %>%
    left_join(centroids, by = "color") %>%
    mutate(
        Distance_to_Centroid = sqrt((PC1 - Mean_PC1)^2 + (PC2 - Mean_PC2)^2)
    ) %>%
    group_by(color) %>%
    slice_min(order_by = Distance_to_Centroid, n = 1, with_ties = FALSE) %>%
    ungroup()
```

Create an image snapshot of each plot for the figure
```{r}

snapshot_dir <- here("3_structure_analysis/plot_snapshots_fig4")
dir.create(snapshot_dir, showWarnings = FALSE, recursive = TRUE)

plot_data$image_path <- NA_character_

all_coords <- as.data.frame(centroids(plots), geom = "XY")[, c("x", "y")]

rows_to_snapshot <- plot_centroids$row_id

for (plot_idx in rows_to_snapshot) {

    plot_id <- plot_data$plot[plot_idx]

    img_fname <- file.path(snapshot_dir, paste0("plot_", plot_id, ".png"))
    plot_data$image_path[plot_idx] <- img_fname

    px <- all_coords[plot_idx, "x"]
    py <- all_coords[plot_idx, "y"]

    roi_i <- clip_circle(nctg, x = px, y = py, radius = 50)

    clear3d()
    rgl::close3d()

    plot(roi_i, color = "RGB", bg = "white", size = 2, axes = FALSE)
    view3d(theta = 0, phi = -45, zoom = 0.5, fov = 120)

    snapshot3d(img_fname, width = 512, height = 512)

    img <- image_read(img_fname)
    img <- image_transparent(img, "white", fuzz = 10)
    image_write(img, img_fname)
}
```

Attach the image paths to the plot_centroids data frame for plotting
```{r}
plot_centroids <- plot_centroids %>%
    mutate(
        image_path = file.path(snapshot_dir, paste0("plot_", plot, ".png"))
    )
```

Plot images of the closest plots to the centroids for each group in PCA space
```{r}

old_PF <- filter(plot_centroids, plot_centroids$color == "old_Pi")

p1 <- ggplot(old_PF, aes(x = 1, y = 1)) +
    geom_image(aes(image = image_path), size = 1.15) +
    scale_x_discrete(position = "top") +
    xlab("Plot 49 - Canary pine forest (old substrate)") +
    theme_minimal() +
    theme(
        legend.position = "none",
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line.x = element_blank(),
        axis.line.y = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 14, colour = "black", hjust = 0),
        axis.ticks.length = element_blank(),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank()
    )

young_PF <- filter(plot_centroids, plot_centroids$color == "young_Pi")

p2 <- ggplot(young_PF, aes(x = 1, y = 1)) +
    geom_image(aes(image = image_path), size = 1.15) +
    scale_x_discrete(position = "top") +
    xlab("Plot 31 - Canary pine forest (young substrate)") +
    theme_minimal() +
    theme(
        legend.position = "none",
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line.x = element_blank(),
        axis.line.y = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 14, colour = "black", hjust = 0),
        axis.ticks.length = element_blank(),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank()
    )

old_FB <- filter(plot_centroids, plot_centroids$color == "old_Fa")

p3 <- ggplot(old_FB, aes(x = 1, y = 1)) +
    geom_image(aes(image = image_path), size = 1.15) +
    scale_x_discrete(position = "top") +
    xlab("Plot 78 - Fayal-Brezal (old substrate)") +
    theme_minimal() +
    theme(
        legend.position = "none",
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line.x = element_blank(),
        axis.line.y = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 14, colour = "black", hjust = 0),
        axis.ticks.length = element_blank(),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank()
    )

young_FB <- filter(plot_centroids, plot_centroids$color == "young_Fa")

p4 <- ggplot(young_FB, aes(x = 1, y = 1)) +
    geom_image(aes(image = image_path), size = 1.15) +
    scale_x_discrete(position = "top") +
    xlab("Plot 87 - Fayal-Brezal (young substrate)") +
    theme_minimal() +
    theme(
        legend.position = "none",
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line.x = element_blank(),
        axis.line.y = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_text(size = 14, colour = "black", hjust = 0),
        axis.ticks.length = element_blank(),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank()
    )

```

Plot the average leaf area density (LAD) profiles for each ecosystem type and substrate age
```{r}

lad_avg_profiles <- LAD_summary %>%
    mutate(color = plot_data$color) %>%
    select(color, ends_with("_mean")) %>%
    pivot_longer(cols = -color, names_to = "layer_name", values_to = "LAD") %>%
    group_by(color, layer_name) %>%
    summarise(
        mean_LAD = mean(LAD, na.rm = TRUE),
        se_LAD = sd(LAD, na.rm = TRUE) / sqrt(n()),
        .groups = "drop"
    ) %>%
    arrange(color, as.numeric(gsub("[^0-9.]", "", layer_name))) %>%
    group_by(color) %>%
    mutate(height = seq(from = 1.75, to = 49.25, by = 2.5)) %>%
    ungroup()

lad_avg_profiles$color <- factor(lad_avg_profiles$color, levels = c("old_Fa", "young_Fa", "old_Pi", "young_Pi"))

h <- ggplot(lad_avg_profiles, aes(x = mean_LAD, y = height, fill = color)) +
    geom_col(orientation = "y", alpha = 0.7) +
    geom_errorbarh(aes(xmin = pmax(0, mean_LAD - se_LAD * 2), xmax = mean_LAD + se_LAD * 2), height = 0.5) +
    scale_fill_manual(values = c("#33658A", "#86BBD8", "#c97704", "#FFC15E")) +
    scale_y_continuous(breaks = c(0, 10, 20, 30, 40), labels = c("0", "10", "20", "30", "40")) +
    scale_x_continuous(breaks = c(0, 0.25, 0.5), labels = c("0", "0.25", "0.5"), limits = c(0, 0.7)) +
    facet_wrap(~color, nrow = 1, ncol = 4) +
    labs(x = "Mean leaf area density [m²/m³]", y = "Height [m]") +
    theme_minimal() +
    theme(
        legend.position = "none",
        strip.text = element_blank(),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line.x = element_line(colour = "black"),
        axis.line.y = element_line(colour = "black"),
        axis.text.x = element_text(size = 12, colour = "black"),
        axis.text.y = element_text(size = 12, colour = "black"),
        axis.title.x = element_text(size = 14, colour = "black"),
        axis.title.y = element_text(size = 14, colour = "black"),
        axis.ticks.length = unit(0.1, "cm"),
        axis.ticks.x = element_line(colour = "black"),
        axis.ticks.y = element_line(colour = "black")
    )

h
```

Plot the hierarchical partitioning results for the structure data
```{r}
res_hier_par_structure <- read_xlsx(here("3_structure_analysis/data/Res_hier_par_structure.xlsx"))

predictors_endemicity <- c(
    "Age" = "#ca0020",
    "Elevation" = "#f4a582",
    "Precipitation" = "#92c5de",
    "Temperature" = "#0571b0"
)

eco_labels <- c(
    Fa = "Fayal-Brezal",
    Pi = "Canary pine forest"
)

# plot it
fs <- 12
hier_par_plot_structure <- ggplot(
    data = res_hier_par_structure, aes(x = Ecosystem, y = Individual * 100, fill = Predictor)
) +
    geom_bar(
        stat = "identity",
        color = "black"
    ) +
    scale_x_discrete(labels = eco_labels) +
    scale_fill_manual(name = NULL, values = predictors_endemicity) +
    scale_y_continuous(limits = c(0, 40)) +
    theme_minimal() +
    theme(
        panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line.x = element_line(colour = "black"),
        axis.line.y = element_line(colour = "black"),
        axis.text.x = element_text(size = 12, colour = "black"),
        axis.text.y = element_text(size = 12, colour = "black"),
        axis.title.x = element_text(size = 14, colour = "black"),
        axis.title.y = element_text(size = 14, colour = "black"),
        axis.ticks.length = unit(0.1, "cm"),
        axis.ticks.x = element_line(colour = "black"),
        axis.ticks.y = element_line(colour = "black")
    ) +
    theme(
        legend.position = c(0.8, 0.8),
        legend.text = element_text(size = 12, colour = "black")
    ) +
    xlab("") +
    ylab("% of explained variance") +
    annotate("text",
        x = 1, y = 40,
        label = "c) Relative importance of predictors",
        hjust = 0,
        size = fs / 2
    ) +
    labs(fill = NULL, pattern = NULL)
hier_par_plot_structure

```

Create compound figures
```{r}

layout_images <- "
AC
BD
"

layout <- "
AAAAAACCCCCCC
AAAAAACCCCCCC
AAAAAACCCCCCC
AAAAAACCCCCCC
AAAAAACCCCCCC
AAAAAACCCCCCC
BBBBBBDDDDDDD
BBBBBBDDDDDDD
BBBBBBDDDDDDD
BBBBBBDDDDDDD
"


plot_images <- p1 + p2 + p3 + p4 +
    plot_layout(design = layout_images)


plot <- p + h + plot_images + hier_par_plot_structure +
    plot_layout(design = layout, axis_titles = "keep")
plot

ggsave(plot = plot, filename = here("figures/Fig_4.pdf"), width = 14, height = 10.5, device = "pdf", dpi = 500)
```


